---
title: "wanting_35507071_TaskD"
author: "Wanting Echo Zhao"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
# Set global chunk options for reproducibility and clean output
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```


# Question 1

## **Mind map**

- Find differences between faculties and sexes: extract these columns and convert to factors.
- Extract them to individual datasets for further testing.
- Compare differences between groups using t-test and ANOVA.
```{r}
# Set working directory to where the data files are located (run this chunk if in VSCode not RStudio project)
setwd("/Users/ez_us/Documents/5145/Assignment4/TaskD")
```
```{r load-and-prepare-data}
# Load data and convert relevant columns to factors
independent_data <- read.csv("TaskD_Dataset/forum_ind_train.csv", stringsAsFactors = FALSE)
independent_data$demographic_sex <- factor(independent_data$demographic_sex)
independent_data$unit_faculty <- factor(independent_data$unit_faculty)

# Preview the data
head(independent_data$demographic_sex)
head(independent_data$unit_faculty)
dplyr::count(independent_data, demographic_sex)

# Identify LIWC variables (all except ID and categoricals)
liwc_vars <- setdiff(names(independent_data), c("Unique_ID", "unit_faculty", "demographic_sex"))
```

### Compare by Sex
```{r compare-by-sex}
# Remove rows with ambiguous or missing sex values
sex_data <- independent_data[independent_data$demographic_sex %in% c("F", "M"), ]
sex_data$demographic_sex <- droplevels(sex_data$demographic_sex)

# Function to run t-test for each LIWC variable by sex
sex_ttest_results <- lapply(liwc_vars, function(v) {
  tt <- t.test(sex_data[[v]] ~ sex_data$demographic_sex)
  data.frame(
    variable = v,
    mean_F = tt$estimate[1],
    mean_M = tt$estimate[2],
    diff = tt$estimate[1] - tt$estimate[2],
    p_value = tt$p.value
  )
})
sex_results <- do.call(rbind, sex_ttest_results) # combine two data frames
sex_results$p_adj <- p.adjust(sex_results$p_value, method = "BH") # Adjust for multiple comparisons to control the false discovery rate or family-wise error rate. Benjamini-Hochberg, controls FDR
sex_results <- sex_results[order(sex_results$p_adj), ]
head(sex_results, 10)
```

### Compare by Faculty
```{r compare-by-faculty}
# Remove rows with missing faculty
faculty_data <- independent_data[!is.na(independent_data$unit_faculty), ]
faculty_data$unit_faculty <- droplevels(faculty_data$unit_faculty)

# Function to run ANOVA for each LIWC variable by faculty
faculty_anova_results <- lapply(liwc_vars, function(v) {
  fit <- aov(faculty_data[[v]] ~ faculty_data$unit_faculty)
  p <- summary(fit)[[1]][["Pr(>F)"]][1]
  data.frame(variable = v, p_value = p)
})
fac_results <- do.call(rbind, faculty_anova_results)
fac_results$p_adj <- p.adjust(fac_results$p_value, method = "BH")
fac_results <- fac_results[order(fac_results$p_adj), ]
head(fac_results, 10)
```

### Group Means for Standout Variable
```{r group-means}
# Example: mean of 'money' variable by faculty
tapply(faculty_data$money, faculty_data$unit_faculty, mean)
```

### Plots
```{r plots}
library(ggplot2)
# Boxplot: Social words by sex
ggplot(sex_data, aes(demographic_sex, social, fill = demographic_sex)) +
  geom_boxplot() +
  labs(title = "Social words by sex", x = "Sex", y = "Social word count") +
  theme_minimal()

# Boxplot: Money words by faculty
ggplot(faculty_data, aes(unit_faculty, money)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Money words by faculty", x = NULL, y = "Money word count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

ggplot(fac_df, aes(unit_faculty, money)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Money words by faculty", x = NULL, y = "money") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
**Sex:**
15 of 93 variables differ significantly (adjusted p < 0.05). The biggest gaps: women's posts score higher on social (8.99 vs 7.84), family (0.19 vs 0.06), Exclam (1.39 vs 0.43), Clout, and are longer overall (WC 122 vs 106 words); men score higher on number (3.47 vs 2.53). These line up with well-documented LIWC gender patterns — more affiliative/social language from women, more numeric language from men.

**Faculty:** 
78 of 93 variables differ — faculty effects are far stronger than sex effects. The standouts are topic words: money is dominated by Business & Economics (mean 4.38 vs ~0.1–0.9 everywhere else), and health/bio/work follow the same disciplinary logic.

Differences between faculties are much larger and more pervasive than differences between sexes, and the strongest faculty differences are content/topic words that reflect what each discipline talks about.

<br>
---


# Question 2

The ind and dep files are joined on Unique_ID to form the modelling table, and a single cleaning function is applied to both train and validation so they are treated identically. I evaluate three classifiers covered in the unit — a decision tree, logistic regression and naive Bayes — each trained on all independent variables. Performance is measured with accuracy, precision, recall and F1, treating Content-relevant as the positive class so that F1 is consistent with the metric used for Q4 and Q5.

library(e1071)
library(MLmetrics)

```{r data-prep-and-modeling}
# Load required libraries
library(rpart)
library(e1071)
library(MLmetrics)

# Helper function: join independent and dependent data by Unique_ID
read_join <- function(ind_file, dep_file) {
  x <- read.csv(ind_file, stringsAsFactors = FALSE)
  y <- read.csv(dep_file, stringsAsFactors = FALSE)
  merge(x, y, by = "Unique_ID")
}

# Helper function: clean dataset (remove duplicates, handle NAs, set factors)
clean <- function(df) {
  df <- df[!duplicated(df$Unique_ID), ]  # Remove duplicate IDs
  num <- setdiff(names(df), c("Unique_ID","unit_faculty","demographic_sex","label"))
  df <- df[rowSums(is.na(df[num])) < length(num), ]  # Remove all-NA rows
  df <- df[!is.na(df$label), ]  # Remove rows with missing label
  df$unit_faculty <- ifelse(is.na(df$unit_faculty), "Unknown", df$unit_faculty)
  df$demographic_sex <- ifelse(df$demographic_sex %in% c("F","M"), df$demographic_sex, "Unknown")
  df$unit_faculty    <- factor(df$unit_faculty)
  df$demographic_sex <- factor(df$demographic_sex)
  df$label <- factor(df$label, levels = c("Content-irrelevant","Content-relevant"))
  df
}

# Prepare training and validation sets
train <- clean(read_join("TaskD_Dataset/forum_ind_train.csv", "TaskD_Dataset/forum_dep_train.csv"))
val   <- clean(read_join("TaskD_Dataset/forum_ind_validation.csv", "TaskD_Dataset/forum_dep_validation.csv"))
val$unit_faculty    <- factor(val$unit_faculty, levels = levels(train$unit_faculty))
val$demographic_sex <- factor(val$demographic_sex, levels = levels(train$demographic_sex))

# Define predictors and formula
predictors <- setdiff(names(train), c("Unique_ID","label"))
form <- as.formula(paste("label ~", paste(predictors, collapse = " + ")))

# Helper function: evaluate model metrics
eval_metrics <- function(actual, predicted, positive = "Content-relevant") {
  data.frame(
    accuracy  = Accuracy(y_pred = predicted, y_true = actual),
    precision = Precision(y_true = actual, y_pred = predicted, positive = positive),
    recall    = Recall(y_true = actual, y_pred = predicted, positive = positive),
    f1        = F1_Score(y_true = actual, y_pred = predicted, positive = positive)
  )
}

# Fit and predict: Decision Tree
fit_tree <- rpart(form, data = train, method = "class")
pred_tree <- predict(fit_tree, val, type = "class")

# Fit and predict: Logistic Regression
fit_glm <- glm(form, data = train, family = binomial)
pred_glm <- ifelse(predict(fit_glm, val, type="response") > 0.5, "Content-relevant", "Content-irrelevant")

# Fit and predict: Naive Bayes
fit_nb <- naiveBayes(form, data = train)
pred_nb <- predict(fit_nb, val)

# Combine results for easy comparison
results <- rbind(
  cbind(model="Decision tree",       eval_metrics(val$label, pred_tree)),
  cbind(model="Logistic regression", eval_metrics(val$label, pred_glm)),
  cbind(model="Naive Bayes",         eval_metrics(val$label, pred_nb))
)
results
```


**Logistic regression** performs best on the validation set *(F1 = 0.917, accuracy = 0.881)*, ahead of the **decision tree** *(F1 = 0.902)* and **naive Bayes** *(F1 = 0.872)*. Logistic regression is chosen as Model 1. 
**Naive Bayes** has the weakest precision, likely due to its strong independence assumption being violated by correlated LIWC variables.

> Note: Fitting logistic regression on all 95 predictors triggers a "fitted probabilities numerically 0 or 1" warning, indicating quasi-separation and a risk of overfitting. This motivates the variable-selection and alternative-model experiments in Q4.


```{r variable-importance-and-univariate-tests}
# Q3: Variable importance and univariate tests
library(rpart)
library(ggplot2)

# Method 1: Decision-tree variable importance
fit_tree <- rpart(form, data = train, method = "class")
imp <- sort(fit_tree$variable.importance, decreasing = TRUE)
head(imp, 15) # Show top 15 important variables

# Plot top 15 important variables
imp_df <- data.frame(variable = names(imp), importance = as.numeric(imp))[1:15, ]
imp_df$variable <- factor(imp_df$variable, levels = rev(imp_df$variable))
ggplot(imp_df, aes(variable, importance)) +
  geom_col() +
  coord_flip() +
  labs(title = "Decision-tree variable importance (top 15)", x = NULL, y = "Importance") +
  theme_minimal()

# Method 2: Univariate t-test for each LIWC variable by label
liwc_vars <- setdiff(predictors, c("unit_faculty","demographic_sex"))

# Split the for-loop for clarity: run t-tests, then build the data frame
t_test_results <- lapply(liwc_vars, function(v) {
  tt <- t.test(train[[v]] ~ train$label)
  data.frame(variable = v,
             mean_irrel = tt$estimate[1],
             mean_rel = tt$estimate[2],
             t_stat = unname(tt$statistic),
             p_value = tt$p.value)
})
uni <- do.call(rbind, t_test_results)
uni$p_adj <- p.adjust(uni$p_value, method = "BH")
uni <- uni[order(abs(uni$t_stat), decreasing = TRUE), ]
head(uni, 15)

# Compare overlap between top variables from both methods
intersect(names(head(imp, 15)), head(uni$variable, 15))

# Categorical predictors: chi-square tests
chisq.test(table(train$demographic_sex, train$label))   # Sex vs label
chisq.test(table(train$unit_faculty, train$label))      # Faculty vs label

# Proportion tables for interpretation
prop.table(table(train$demographic_sex, train$label), 1) # Relevance rate by sex
prop.table(table(train$unit_faculty, train$label), 1)    # Relevance rate by faculty
```


Two independent methods — decision-tree importance and a univariate t-test of each variable by class — agree on a core set of important variables (WC, WPS, cause, i, ppron, pronoun, affiliation). Word count is by far the strongest signal: content-relevant posts are much longer (160 vs 36 words) and more elaborated (WPS, Sixltr, Analytic). 

Conversely, content-irrelevant posts use more first-person and personal pronouns and affiliation words, reflecting short social or logistical messages. In short, relevance is driven by post length and a formal-reasoning-versus-personal-chat contrast. This informs Q4, where I test whether a model built on this important subset can outperform Model 1.

The two categorical variables were tested with chi-square, the appropriate test for categorical predictors. Sex is not associated with the label (χ² = 3.03, p = 0.22; relevance rate ≈ 64% for famele and 62.8% for male), whereas faculty is strongly associated (χ² = 141.8, p < 0.001), with content-relevant rates ranging from about 46% (Engineering, Law) to 80% (Business & Economics, Arts). 

Thus the important predictors are post length and analytic-versus-personal language (numeric LIWC variables) together with the unit's faculty; the author's sex carries no predictive value.

Because here is juding "relevant/irrelevant", utilizing "class" method for `rpart()`(desision tree) functiont to predicts a discrete category or class label.


---

# Question 4
```{r}
library(rpart) 
library(e1071)
library(class)

```{r model-comparison-and-improvement}
# Q4: Model comparison and improvement
library(rpart)
library(e1071)
library(class)
library(MLmetrics)

# 1. Isolate variable types
num_vars <- setdiff(names(train), c("Unique_ID","unit_faculty","demographic_sex","label"))
all_vars <- setdiff(names(train), c("Unique_ID","label"))

# 2. Select top 20 most important variables by t-statistic magnitude
# Split the for-loop: compute t-stats, then select top variables
get_t_stat <- function(variable_name) {
  test_result <- t.test(train[[variable_name]] ~ train$label)
  abs(unname(test_result$statistic))
}
tt_abs <- sapply(num_vars, get_t_stat)
important <- names(sort(tt_abs, decreasing = TRUE))[1:20]

# 3. Log-transform LIWC variables to address skewness
trainL <- train
valL   <- val
trainL[num_vars] <- log1p(train[num_vars])
valL[num_vars]   <- log1p(val[num_vars])

# 4. Initialize model leaderboard
cand <- list()

# Model 1: Baseline Logistic Regression (Raw Data)
```
To improve Model 1, I applied several of the strategies suggested in the question, each motivated by a specific weakness of the baseline. Model 1 (logistic regression on all raw variables) triggered a "fitted probabilities 0 or 1" warning, signalling quasi-separation and overfitting, so I tested: a data transformation (log1p of the LIWC variables, since they are right-skewed counts), an alternative algorithm that handles non-linear structure (a decision tree), and a rescaled distance-based model on a feature subset (kNN on the 20 most important variables, standardised because kNN is scale-sensitive).


# Model 2: Improved Logistic Regression (Log-Transformed Data)
The validation F1 scores were: baseline logistic = 0.917, log-transformed logistic = 0.923, decision tree = 0.902, and kNN (k = 9) = 0.916. The log-transformed logistic regression was the best, and just as importantly it no longer produced the separation warning — log-compressing the extreme counts (e.g. very long posts in WC) yields a better-conditioned, less overfit model. The decision tree and kNN did not beat the baseline.

I therefore conclude that Model 1 can be modestly but genuinely improved: log-transforming the predictors lifts validation F1 from 0.917 to 0.923 while removing the overfitting risk. This log-transformed logistic regression is the model I carry forward to predict the test set in Q5.

# Model 3: Decision Tree
tree_model <- rpart(reformulate(all_vars, "label"), data = train, method = "class")
pred_tree  <- predict(tree_model, val, type = "class")


# Model 4: kNN using only the 20 'important' features
sc    <- scale(train[num_vars])
va_sc <- scale(val[num_vars], center = attr(sc, "scaled:center"), scale = attr(sc, "scaled:scale"))
pred_knn <- knn(train = sc[, important], test = va_sc[, important], cl = train$label, k = 9)


# 5. Print out the leaderboard
data.frame(model = names(cand), F1 = round(unlist(cand), 4))

# 6. Final analysis of the winner (logistic regression on log-transformed data)
F1_Score(y_true = val$label, y_pred = pred_log, positive = "Content-relevant")  # F1 validation score
table(predicted = pred_log, actual = val$label) # Confusion matrix
```
---

# Question 5


```{r test-set-prediction-and-submission}
# Q5: Predict on test set and prepare submission

# Final model: logistic regression on log-transformed variables
num_vars <- setdiff(names(train), c("Unique_ID","unit_faculty","demographic_sex","label"))
all_vars <- setdiff(names(train), c("Unique_ID","label"))

trainL <- train
trainL[num_vars] <- log1p(train[num_vars])
final_model <- glm(reformulate(all_vars, "label"), data = trainL, family = binomial)

# Read the test files (do NOT drop rows: every row needs a label)
ind_test <- read.csv("TaskD_Dataset/forum_ind_test.csv", stringsAsFactors = FALSE)
dep_test <- read.csv("TaskD_Dataset/forum_dep_test.csv", stringsAsFactors = FALSE)

# Clean up categoricals and apply same factor levels as training
ind_test$unit_faculty    <- ifelse(is.na(ind_test$unit_faculty), "Unknown", ind_test$unit_faculty)
ind_test$demographic_sex <- ifelse(ind_test$demographic_sex %in% c("F","M"), ind_test$demographic_sex, "Unknown")
ind_test$unit_faculty    <- factor(ind_test$unit_faculty,    levels = levels(train$unit_faculty))
ind_test$demographic_sex <- factor(ind_test$demographic_sex, levels = levels(train$demographic_sex))
ind_test[num_vars]       <- log1p(ind_test[num_vars])  # Apply same transformation

# Predict and fill dep_test's label column in its original row order
prob <- predict(final_model, ind_test, type = "response")
pred <- ifelse(prob > 0.5, "Content-relevant", "Content-irrelevant")
pred_by_id <- data.frame(Unique_ID = ind_test$Unique_ID, pred = pred)
dep_test$label <- pred_by_id$pred[match(dep_test$Unique_ID, pred_by_id$Unique_ID)]

# Handle any test row with all-NA features by assigning the majority class
dep_test$label[is.na(dep_test$label)] <- names(sort(table(train$label), decreasing = TRUE))[1]

# Write submission: same 2 columns, 556 rows, unquoted
write.csv(dep_test, "Zhao_35507071_forum_label_test.csv", row.names = FALSE, quote = FALSE)
```


