---
title: "wanting_35507071_TaskD"
author: "Wanting Echo Zhao"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```

# Question 1

## **Mind map**

- Finding differences between faculties and sexes. - get these two columns. change it to factor format, instead of numeral data.
- Extract them to a individual data set for further test.
- Think about the ways of comparing difference between multiple groups. - t test, Anova test.


```{r}
# Load data and extract two target columns. set them as factors
ind <- read.csv("TaskD_Dataset/forum_ind_train.csv", stringsAsFactors = F)
ind$demographic_sex <- factor(ind$demographic_sex)
ind$unit_faculty <- factor(ind$unit_faculty)

head(ind$demographic_sex)
head(ind$unit_faculty)
ind |> count(demographic_sex)


# Set the LIWC variables = everything except ID and the 2 categoricals.
# setDiff() can be regarded as left_join
liwc_vars <- setdiff(names(ind), c("Unique_ID", "unit_faculty", "demographic_sex"))

#  ===== SEX ======
# There are only 2 'X',5 'NA', drop them directly.
sex_df <- ind[ind$demographic_sex %in% c("F", "M"), ] 
sex_df$demographic_sex <- droplevels(sex_df$demographic_sex) # droplevel() remove value not 'F' or 'M'

sex_results <- data.frame()
for (v in liwc_vars) {
  # two-sample t-test
  tt <- t.test(sex_df[[v]] ~ sex_df$demographic_sex)  
  # sex_df[[v]] extracts the numerical data for the current language variable v. 
  # ~ sex_df$demographic_sex splits that data into two groups based on sex
  sex_results <- rbind(sex_results, data.frame(
    variable = v,
    mean_F = tt$estimate[1], 
    mean_M = tt$estimate[2],
    diff = tt$estimate[1] - tt$estimate[2], 
    p_value = tt$p.value))
}
sex_results$p_adj <- p.adjust(sex_results$p_value, method = "BH")  # correct for 93 tests
sex_results <- sex_results[order(sex_results$p_adj), ]  # top10 significant
head(sex_results, 10)


# ====== Faculty ======
fac_df <- ind[!is.na(ind$unit_faculty), ]
fac_df$unit_faculty <- droplevels(fac_df$unit_faculty)

fac_results <- data.frame()
for (v in liwc_vars) {
  fit <- aov(fac_df[[v]] ~ fac_df$unit_faculty)  # one-way ANOVA
  p   <- summary(fit)[[1]][["Pr(>F)"]][1]  # pull the p-value out
  fac_results <- rbind(fac_results, data.frame(variable = v, p_value = p))
}
fac_results$p_adj <- p.adjust(fac_results$p_value, method = "BH")
fac_results <- fac_results[order(fac_results$p_adj), ]
head(fac_results, 10)

# Group-means table for the standout faculty variable
tapply(fac_df$money, fac_df$unit_faculty, mean)

# ====== PLOTS ======
library(ggplot2)
ggplot(sex_df, aes(demographic_sex, social, fill = demographic_sex)) +
  geom_boxplot() + labs(title = "Social words by sex", x = "Sex", y = "social") +
  theme_minimal()

ggplot(fac_df, aes(unit_faculty, money)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Money words by faculty", x = NULL, y = "money") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
**Sex:**
15 of 93 variables differ significantly (adjusted p < 0.05). The biggest gaps: women's posts score higher on social (8.99 vs 7.84), family (0.19 vs 0.06), Exclam (1.39 vs 0.43), Clout, and are longer overall (WC 122 vs 106 words); men score higher on number (3.47 vs 2.53). These line up with well-documented LIWC gender patterns — more affiliative/social language from women, more numeric language from men.

**Faculty:** 
78 of 93 variables differ — faculty effects are far stronger than sex effects. The standouts are topic words: money is dominated by Business & Economics (mean 4.38 vs ~0.1–0.9 everywhere else), and health/bio/work follow the same disciplinary logic.

Differences between faculties are much larger and more pervasive than differences between sexes, and the strongest faculty differences are content/topic words that reflect what each discipline talks about.

<br>
---

# Qustion2

The ind and dep files are joined on Unique_ID to form the modelling table, and a single cleaning function is applied to both train and validation so they are treated identically. I evaluate three classifiers covered in the unit — a decision tree, logistic regression and naive Bayes — each trained on all independent variables. Performance is measured with accuracy, precision, recall and F1, treating Content-relevant as the positive class so that F1 is consistent with the metric used for Q4 and Q5.

```{r}
library(rpart)
library(e1071)
library(MLmetrics)


read_join <- function(ind_file, dep_file) {
  x <- read.csv(ind_file, stringsAsFactors = FALSE)
  y <- read.csv(dep_file, stringsAsFactors = FALSE)
  merge(x, y, by = "Unique_ID")
}

# write a function to clean datasets used here
clean <- function(df) {
  df <- df[!duplicated(df$Unique_ID), ]  # drop duplicate IDs
  num <- setdiff(names(df), c("Unique_ID","unit_faculty","demographic_sex","label"))
  df <- df[rowSums(is.na(df[num])) < length(num), ]  # drop all-NA junk rows
  df <- df[!is.na(df$label), ]  # drop rows with no label
  df$unit_faculty <- ifelse(is.na(df$unit_faculty), "Unknown", df$unit_faculty)
  df$demographic_sex <- ifelse(df$demographic_sex %in% c("F","M"), # As Q1
                               df$demographic_sex, "Unknown") # label 'X' and 'NA' to 'Unknown'
  df$unit_faculty    <- factor(df$unit_faculty)
  df$demographic_sex <- factor(df$demographic_sex)
  df$label <- factor(df$label, levels = c("Content-irrelevant","Content-relevant"))
  df
}

train <- clean(read_join("TaskD_Dataset/forum_ind_train.csv",
                         "TaskD_Dataset/forum_dep_train.csv"))
val <- clean(read_join("TaskD_Dataset/forum_ind_validation.csv",
                         "TaskD_Dataset/forum_dep_validation.csv"))
val$unit_faculty<- factor(val$unit_faculty, levels = levels(train$unit_faculty))
val$demographic_sex <- factor(val$demographic_sex, levels = levels(train$demographic_sex))

predictors <- setdiff(names(train), c("Unique_ID","label"))
form <- as.formula(paste("label ~", paste(predictors, collapse = " + ")))

eval_metrics <- function(actual, predicted, positive = "Content-relevant") {
  data.frame(
    # NOTE: Accuracy uses (y_pred, y_true); the others use (y_true, y_pred).
    # I name the arguments so the order can't be mixed up.
    accuracy  = Accuracy(y_pred = predicted, y_true = actual),
    precision = Precision(y_true = actual, y_pred = predicted, positive = positive),
    recall = Recall(y_true = actual, y_pred = predicted, positive = positive),
    f1 = F1_Score(y_true = actual, y_pred = predicted, positive = positive)
  )
}

fit_tree <- rpart(form, data = train, method = "class")
pred_tree <- predict(fit_tree, val, type = "class")

fit_glm <- glm(form, data = train, family = binomial)
pred_glm <- ifelse(predict(fit_glm, val, type="response") > 0.5,
                   "Content-relevant", "Content-irrelevant")

fit_nb <- naiveBayes(form, data = train)
pred_nb <- predict(fit_nb, val)

rbind(
  cbind(model="Decision tree",       eval_metrics(val$label, pred_tree)),
  cbind(model="Logistic regression", eval_metrics(val$label, pred_glm)),
  cbind(model="Naive Bayes",         eval_metrics(val$label, pred_nb))
)
```
**Logistic regression** performs best on the validation set *(F1 = 0.917, accuracy = 0.881)*, ahead of the **decision tree** *(F1 = 0.902)* and **naive Bayes** *(F1 = 0.872)*. I therefore take logistic regression as Model 1. 
**Naive Bayes** has the weakest precision, which fits its strong independence assumption being violated by the many correlated LIWC variables.

> Note that fitting logistic regression on all 95 predictors triggers a "fitted probabilities numerically 0 or 1" warning, indicating quasi-separation and a risk of overfitting. This motivates the variable-selection and alternative-model experiments in Q4.

```{r}
library(rpart)
library(ggplot2)
# (train built with the same read_join + clean from Q2)

# Method 1: decision-tree importance
# Because here is juding "relevant/irrelevant", utilizing "class" method for `rpart()`(desision tree) functiont to predicts a discrete category or class label.
fit_tree <- rpart(form, data = train, method = "class")
imp <- sort(fit_tree$variable.importance, decreasing = TRUE)
head(imp, 15)

d <- data.frame(variable = names(imp), importance = as.numeric(imp))[1:15, ]
d$variable <- factor(d$variable, levels = rev(d$variable))
ggplot(d, aes(variable, importance)) + 
  geom_col() +
  coord_flip() + 
  labs(
    title = "Decision-tree variable importance (top 15)",
    x = NULL, y = "Importance") + 
  theme_minimal()

# Method 2: apply t-test of each LIWC variable by label
liwc_vars <- setdiff(predictors, c("unit_faculty","demographic_sex"))
uni <- data.frame()

for (v in liwc_vars) {
  tt <- t.test(train[[v]] ~ train$label)
  uni <- rbind(uni, data.frame(variable = v,
              mean_irrel = tt$estimate[1], mean_rel = tt$estimate[2],
              t_stat = unname(tt$statistic), p_value = tt$p.value))
}

uni$p_adj <- p.adjust(uni$p_value, method = "BH")
uni <- uni[order(abs(uni$t_stat), decreasing = TRUE), ]
head(uni, 15)

intersect(names(head(imp,15)), head(uni$variable,15))   # the robust core

# Categorical predictors need chi-square, not a t-test
chisq.test(table(train$demographic_sex, train$label))   
# p = 0.22  -> not associated

chisq.test(table(train$unit_faculty,   train$label))
# p < 2e-16 -> strongly associated

prop.table(table(train$demographic_sex, train$label), 1)
# relecanxe rate of sexes

prop.table(table(train$unit_faculty, train$label), 1)
# relevance rate per faculty
```


Two independent methods — decision-tree importance and a univariate t-test of each variable by class — agree on a core set of important variables (WC, WPS, cause, i, ppron, pronoun, affiliation). Word count is by far the strongest signal: content-relevant posts are much longer (160 vs 36 words) and more elaborated (WPS, Sixltr, Analytic). 

Conversely, content-irrelevant posts use more first-person and personal pronouns and affiliation words, reflecting short social or logistical messages. In short, relevance is driven by post length and a formal-reasoning-versus-personal-chat contrast. This informs Q4, where I test whether a model built on this important subset can outperform Model 1.

The two categorical variables were tested with chi-square, the appropriate test for categorical predictors. Sex is not associated with the label (χ² = 3.03, p = 0.22; relevance rate ≈ 64% for famele and 62.8% for male), whereas faculty is strongly associated (χ² = 141.8, p < 0.001), with content-relevant rates ranging from about 46% (Engineering, Law) to 80% (Business & Economics, Arts). 

Thus the important predictors are post length and analytic-versus-personal language (numeric LIWC variables) together with the unit's faculty; the author's sex carries no predictive value.

Because here is juding "relevant/irrelevant", utilizing "class" method for `rpart()`(desision tree) functiont to predicts a discrete category or class label.


---

# Question 4

```{r}
library(rpart) 
library(e1071)
library(class)
library(MLmetrics)

# 1. Isolate variable types
num_vars <- setdiff(names(train), c("Unique_ID","unit_faculty","demographic_sex","label"))
all_vars <- setdiff(names(train), c("Unique_ID","label"))

# 2. Filter out the top 20 most important variables via t-test magnitude
# (1): Write a function collect t-value for each words 
get_t_stat <- function(variable_name) {
  test_result <- t.test(train[[variable_name]] ~ train$label)
  absolute_t <- abs(unname(test_result$statistic))
  return(absolute_t)
}

# (2): Feed your list of variables and your new function into sapply
tt_abs <- sapply(num_vars, get_t_stat)

# (3): Pick the top 20 just like before
important <- names(sort(tt_abs, decreasing = TRUE))[1:20]

# 3. Create log-transformed copies to fix right-skewed text distributions
trainL <- train
valL   <- val
trainL[num_vars] <- log1p(train[num_vars])
valL[num_vars]   <- log1p(val[num_vars])

# 4. Initialize model baking contest list
cand <- list()

# Model 1: Baseline Logistic Regression (Raw Data)
m_base    <- suppressWarnings(glm(reformulate(all_vars, "label"), data = train, family = binomial))
pred_base <- ifelse(predict(m_base, val, type = "response") > 0.5, "Content-relevant", "Content-irrelevant")
cand[["logistic all (Model 1)"]] <- F1_Score(y_true = val$label, y_pred = pred_base, positive = "Content-relevant")

# Model 2: Improved Logistic Regression (Log-Transformed Data)
m_log    <- glm(reformulate(all_vars, "label"), data = trainL, family = binomial)
pred_log <- ifelse(predict(m_log, valL, type = "response") > 0.5, "Content-relevant", "Content-irrelevant")
cand[["logistic log1p all (improved)"]] <- F1_Score(y_true = val$label, y_pred = pred_log, positive = "Content-relevant")

# Model 3: Decision Tree
tree_model <- rpart(reformulate(all_vars, "label"), data = train, method = "class")
pred_tree  <- predict(tree_model, val, type = "class")
cand[["decision tree all"]] <- F1_Score(y_true = val$label, y_pred = pred_tree, positive = "Content-relevant")

# Model 4: kNN using only the 20 'important' features
sc    <- scale(train[num_vars])
va_sc <- scale(val[num_vars], center = attr(sc, "scaled:center"), scale = attr(sc, "scaled:scale"))
pred_knn <- knn(train = sc[, important], test = va_sc[, important], cl = train$label, k = 9)
cand[["kNN important k=9"]] <- F1_Score(y_true = val$label, y_pred = pred_knn, positive = "Content-relevant")

# 5. Print out the leaderboard scoreboard
data.frame(model = names(cand), F1 = round(unlist(cand), 4))


# 6. FINAL ANALYSIS OF THE WINNER (Logistic on Log-Transformed Data)
# Yields the F1 validation score: 0.9227
F1_Score(y_true = val$label, y_pred = pred_log, positive = "Content-relevant")  

# Print final Confusion Matrix
table(predicted = pred_log, actual = val$label)
```
To improve Model 1, I applied several of the strategies suggested in the question, each motivated by a specific weakness of the baseline. Model 1 (logistic regression on all raw variables) triggered a "fitted probabilities 0 or 1" warning, signalling quasi-separation and overfitting, so I tested: a data transformation (log1p of the LIWC variables, since they are right-skewed counts), an alternative algorithm that handles non-linear structure (a decision tree), and a rescaled distance-based model on a feature subset (kNN on the 20 most important variables, standardised because kNN is scale-sensitive).

The validation F1 scores were: baseline logistic = 0.917, log-transformed logistic = 0.923, decision tree = 0.902, and kNN (k = 9) = 0.916. The log-transformed logistic regression was the best, and just as importantly it no longer produced the separation warning — log-compressing the extreme counts (e.g. very long posts in WC) yields a better-conditioned, less overfit model. The decision tree and kNN did not beat the baseline.

I therefore conclude that Model 1 can be modestly but genuinely improved: log-transforming the predictors lifts validation F1 from 0.917 to 0.923 while removing the overfitting risk. This log-transformed logistic regression is the model I carry forward to predict the test set in Q5.


---

# Question 5

```{r}
# Final model = the Q4 winner: logistic regression on log-transformed variables
num_vars <- setdiff(names(train), c("Unique_ID","unit_faculty","demographic_sex","label"))
all_vars <- setdiff(names(train), c("Unique_ID","label"))

trainL <- train
trainL[num_vars] <- log1p(train[num_vars])
final_model <- glm(reformulate(all_vars, "label"), data = trainL, family = binomial)

# --- read the test files (do NOT drop rows: every row needs a label) ---
ind_test <- read.csv("TaskD_Dataset/forum_ind_test.csv", stringsAsFactors = FALSE)
dep_test <- read.csv("TaskD_Dataset/forum_dep_test.csv", stringsAsFactors = FALSE)

# same categorical clean-up + identical factor levels as training
ind_test$unit_faculty    <- ifelse(is.na(ind_test$unit_faculty), "Unknown", ind_test$unit_faculty)
ind_test$demographic_sex <- ifelse(ind_test$demographic_sex %in% c("F","M"),
                                   ind_test$demographic_sex, "Unknown")
ind_test$unit_faculty    <- factor(ind_test$unit_faculty,    levels = levels(train$unit_faculty))
ind_test$demographic_sex <- factor(ind_test$demographic_sex, levels = levels(train$demographic_sex))
ind_test[num_vars]       <- log1p(ind_test[num_vars])          # same transform as the model

# predict, then fill dep_test's label column in its ORIGINAL row order
prob <- predict(final_model, ind_test, type = "response")
pred <- ifelse(prob > 0.5, "Content-relevant", "Content-irrelevant")
pred_by_id <- data.frame(Unique_ID = ind_test$Unique_ID, pred = pred)
dep_test$label <- pred_by_id$pred[match(dep_test$Unique_ID, pred_by_id$Unique_ID)]

# one test row has all-NA features -> fall back to the majority class
dep_test$label[is.na(dep_test$label)] <- names(sort(table(train$label), decreasing = TRUE))[1]

# write submission: same 2 columns, 556 rows, unquoted
write.csv(dep_test, "Zhao_35507071_forum_label_test.csv", row.names = FALSE, quote = FALSE)

```


